Thesis: draft

### 2021-05-30 draft.tex
latest update

Testing the opinion dynamics model on network models, gives us useful insights in the interplay between network topologies and model features such as filtering algorithm and stubbornness. It is, however, interesting to test the model on real-world networks to get a better understanding of real-world opinion dynamics.\\
\newline
The results of the two real-world networks: (i) the Last.fm on-line social network; and (ii) the PGP trust network, discussed in Section \ref{realWorld}, are shared and discussed. For each real-world network, the results are compared to results from network models with similar average degree, clustering or modularity. We focus on the formation of echo chambers. We study balanced starting conditions and we introduce a fraction of completely stubborn actors. For both real-world networks, the REC and PR filtering algorithms are compared.

@@

The formation of echo chambers versus the fraction of stubborn actors, $f_s$, in the Last.fm on-line social network is studied. Results are compared to results from network models, all with a size of $N = 8003$ nodes: (i) WS model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM with 53 communities of size 151 nodes each, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) SBM-WS with 53 communities of size 151 nodes each, each community has WS parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithms are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The SBM-WS has similar clustering and modularity as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering or modularity that drives the formation of echo chambers. Other topologies must play a role as well. The most significant difference between the real-world Last.fm network and the SBM-WS lies in the degree distribution: the Last.fm network has a heterogeneous instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to network models with similar clustering and modularity, but with a homogeneous degree distribution.\\ % add parameters of SBM-WS power??
\newline
It is interesting to notice that the Last.fm on-line social network coincides with the SBM with similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of clustering.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The Last.fm on-line social network is compared to network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The Last.fm on-line social network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line represents a constant amount of echo chambers over time. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\newpage
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease. The increase is the strongest for network models with high clustering (WS, SBM-WS). Even though the Last.fm network has similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. The results for the Last.fm network coincide with those of the SBM. This seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering. \\
\newline
Another difference between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The SBM and SBM-WS only have communities of a single, fixed size. This does not play a significant role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_REC.png}
    \label{pr_pow_single}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_PR.png}
    \label{rec_pow_single}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. The SBM-WS with heterogeneous community sizes is compared to the SBM-WS with a single community size. Results are compared to the real-world Last.fm network with similar clustering and modularity. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. The SBM-WS with heterogeneous community sizes (SBM-WS1) is compared to the SBM-WS with a single community size (SBM-WS2). Results are compared to the real-world Last.fm network with similar clustering and modularity. The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line represents a constant amount of echo chambers over time. Each curve is the average of $50$ independent simulations. Error bars are omitted for improved visualization.}
\label{power_vs_single}
\end{figure}

@@

Results for the PGP trust network are compared to results for network models with similar average degree, clustering or modularity. These models are, all with a size of $N = 10680$ nodes: (i) WS model with $\left<k\right> = 4$, $\beta = 0.2$ and $\left<cc\right> = 0.266 \pm 0.003$; (ii) SBM with 60 communities of size 178 nodes each, $p_\text{cl} = 0.02$, $p_\text{add} = 0.00005$, $\left<k\right> = 4.06 \pm 0.02$ and $Q = 0.868 \pm 0.002$; (iii) SBM-WS with 60 communities of size 178 nodes each, each community has WS parameters: $\left<k\right> = 4$, $\beta = 0.12$; edges are added between communities with probability $p_\text{add} = 0.00007$, $\left<k\right> = 4.74 \pm 0.01$, $\left<cc\right> = 0.263 \pm 0.003$ and $Q = 0.842 \pm 0.002$.\\
\newline
In Fig. \ref{echo_vs_fracRes_real_PGP} the formation of echo chambers versus the fraction of stubborn actors, $f_s$, is studied for the real-world PGP trust network and for the network models. This is done for both the PR and REC filtering algorithms. \\
\newline
The PR filtering algorithm (Fig. \ref{pr_real_PGP}) shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors for both the PGP trust network and the network models. The real-world PGP network has a significantly smaller formation of echo chambers compared to the network model with similar clustering and modularity (SBM-WS). This is most likely due to the heterogeneous degree pattern of the PGP network.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_PGP_8x7.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_PGP_8x7.png}
    \label{rec_real_PGP}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The PGP trust network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The PGP trust network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line represents a constant amount of echo chambers over time. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real_PGP}
\end{figure}
\newpage
\noindent
The REC filtering algorithm (Fig. \ref{rec_real_PGP}) shows an increase in the formation of echo chambers with an increasing fraction of stubborn actors, followed by a decrease if the fraction of stubborn actors is further increased. This is visible for both the PGP trust network and the network models. The PGP network has a smaller formation of echo chambers than the network model with similar clustering and modularity (SBM-WS), due to its heterogeneous degree distribution. The data points for the SBM and the PGP trust network coincide, which may suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering.

@@

We made use of simple toy models of social media platforms to generate results regarding the evolution of the prevalence of the opinions and formation of echo chambers. These toy models consist of several layers, including a network layer and an opinion dynamics layer. Filtering algorithms were implemented to replicate and simulate the content curation used by social media companies. Furthermore, stubbornness was introduced to study the interplay between individual resistance to change, network structure and filtering algorithm. For simplicity we only allowed two competing opinions (A and B). We studied two starting conditions: balanced and unbalanced starting conditions.\\

@@

When introducing stubbornness, we found that the interplay between the PR filtering algorithm and individual resistance resulted in a decrease in the formation of echo chambers for the both the high clustering (WS, SBM-WS) and the high modularity networks (SBM, SBM-WS). The REF and REC filtering algorithms showed an increase in the formation of echo chambers with increasing stubbornness, followed by a decrease if stubbornness was further increased. This is mainly seen in networks with high clustering, however a small bump can also be observed in networks with high modularity. This interplay between the PR filtering algorithm and individual resistance versus the interplay between the REF and REC filtering algorithms and stubbornness is seen for both balanced and unbalanced starting conditions. Furthermore, it was found that including stubborn actors hampers the increase in the majority opinion (and thus prevents the formation of a final consensus state) in case of unbalanced starting conditions combined with the PR filtering algorithm.\\

@@

These results give us insight in the interplay between network structure, filtering algorithms and individual resistance. These insights may be crucial to understand phenomena such as polarization in society and to determine the impact of social media on these phenomena. They learn us that even though semantic filtering (PR) enhances the formation of echo chambers in the absence of stubborn actors, this enhancement does not stand on its own. An interplay between networks with high clustering or community structure is needed to effectively see the increase in echo chambers induced by semantic filtering. Random networks, combined with semantic filtering, do not enhance the formation of echo chambers. If there are no stubborn actors, recent or random filtering (REC or REF) seem to be more protective against the formation of echo chambers than semantic filtering. If there is a small fraction of stubborn actors, this scenario changes: semantic filtering starts to hamper the formation of echo chambers, whereas recent or random filtering begins to enhance polarization. It is often assumed that algorithmic personalization, tailored to ones preferences (e.g. semantic filtering), is one of the reasons for the appearance of echo chambers on social media \cite{Ge2020}\cite{Mohseni2018}\cite{Stark2020}. Our results may, however, suggest, that if there is stubbornness (which is probably the case in real life), heterogeneous exposure (REF or REC) to different opinions may be more harmful than homogeneous exposure (PR). Bail et al. found experimental results that may agree with this finding \cite{Bail2018}. This is a result that should be studied more thoroughly.\\

@@

It was found that the initial distribution of the opinions over the communities in networks with community structure (SBM) makes a difference. We investigated a scenario in which a fraction of the communities was given a single opinion, opinion A, while the remaining communities were given an initial $50/50$ opinion distribution. This scenario was then compared to the scenario in which the total opinion distribution was randomly distributed. This was done for both a high and low modularity configuration and for both the REC and PR filtering algorithms. \\

@@

The REC filtering algorithm did not show a difference in the evolution of the prevalence of the opinions between the (initial) community distributed and (initial) randomly distributed scenario. For the PR filtering algorithm, it was found that the high modularity configuration hampers the increase in the majority opinion in the community distributed scenario compared to the random distributed scenario. For the low modularity network, such a discrepancy was not found. Furthermore, it was found that, when using the REC filtering algorithm, the system evolves towards a configuration with more mixed opinion neighborhoods in the community distributed scenario. This reduces polarization and was found for both the high and low modularity configuration. The randomly distributed scenario does not evolve towards mixed opinion neighborhoods, instead a small increase in the formation of echo chambers around the minority opinion B is observed. Hence, polarization increases in this scenario.\\

@@

For the PR filtering algorithm, the random distributed scenario resulted in an increase in the formation of echo chambers around the majority opinion A (for both the low and high modularity configuration). The community distributed scenario for the low modularity configuration hampers this increase. It was found that this is more than just a result of the increased number of initial opinion A echo chambers. For the high modularity network, the community distributed scenario showed an even larger hampering of the formation of echo chambers around the majority opinion A. This is probably the result of: (i) an even higher initial number of echo chambers around the majority opinion A than for the low modularity network; (ii) the hampering in the increase of the majority opinion A for the community distributed scenario compared to the randomly distributed scenario. \\

@@

Finally, results for real-world networks were compared to those from network models. It was found that the formation of echo chambers in the real-world networks is smaller than those in network models with similar clustering and modularity. The reason for this hampered formation of echo chambers was attributed to the heterogeneous degree pattern of the real-world networks. This hampering of echo chambers due to a heterogeneous degree distribution was also found by Perra and Rocha \cite{Perra2019} when they compared network models with heterogeneous degree distribution to models with homogeneous contact patterns.\\

@@

The hampering in the increase of the majority opinion in case of unbalanced starting conditions combined with the introduction of stubborn actors for the PR filtering algorithm agrees partly with related research on continuous opinion dynamics models combined with stubborn actors \cite{Como2013}. In this research, Acemo\u{g}lu et al. showed that introducing stubborn actors (with different opinions) in an inhomogeneous stochastic gossip process of continuous opinion dynamics leads to a situation in which consensus can no longer be reached \cite{Como2013}. This result is also found in the majority opinion dynamics model and in the voter model combined with stubborn actors (equally distributed over the two opinions) \cite{Galam2007}\cite{Mobilia2007}\cite{Yildiz2013}. In our case, the possible consensus around the majority opinion, that could have been reached when combining the PR filtering algorithm with unbalanced starting conditions, can no longer be obtained when introducing stubborn actors.\\

@@

It must be emphasized that there are limitations to our model and methods. Real-life social behavior and opinion dynamics on social media may be far from simple. As a consequence, there may be other dynamics of real-life opinion formation that are not well-captured by our model. More empirical data is therefore needed to compare the results, obtained from our model, against \cite{Castellano2009}\cite{Pawel}. Other limitations may be the small system sizes that were used in our simulations ($N = 10^3$). It is unclear how system size affects the results and it is recommended that the model is tested on both larger network models and real-world networks. Limitations due to the activation mechanism, may be that we used a constant activation probability. Analyses of real social networks show that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. It might be an interesting extension to switch to the heterogeneous activation probabilities and investigate possible impacts on the dynamics. Another possible shortcoming is the binary nature of the model. There are real-life cases in which people have to choose between two competing opinions. These situations, however, do not cover the complete range of opinion dynamics situations. It would be interesting to run the model on configurations where more than two opinions are allowed or to switch to continuous opinion dynamics. We also did not differentiate between the actors in our model, except for non-stubborns versus stubborns. This combined with the homogeneous degree distribution of our network models may be another limitation. Real-life social media platforms display a heterogeneous contact pattern, where the `hubs' are often more influential than other users, not only by their larger reach, but also by their status. This may impact the dynamics of opinions \cite{Fan2020}\cite{Wang2019}. We could adapt our model to mimic this social media behavior. Therefore, a heterogeneous network model should be introduced, combined with an influence parameter. This parameter is used to determine the influence of an actor on others: the opinion of a more influential actor is easier adopted by the others. In the model, the high degree nodes should then be made more influential. The fixed stubbornness of the actors might be another shortcoming. Even though the stubbornness parameter $r$ could be individualized to each actor, we did not include the possibility to be more stubborn towards opinions coming from one friend than from another. This idea may be linked to the influence of certain friends: somebody may adopt another viewpoint easier from certain friends than from others. As a final limitation, we mention the homogeneous degree distribution of our network models. Real-world social networks often have heterogeneous contact patterns. It is therefore recommended to test the opinion dynamics model on heterogeneous network models.\\

>>>>

Studying the opinion dynamics model on network models, gives us insights in the interplay between network topologies and model features such as filtering algorithm and stubbornness. It is, however, interesting to study the model on real-world networks to get a better understanding of real-world opinion dynamics.\\
\newline
The results of the two real-world networks: (i) the Last.fm on-line social network; and (ii) the PGP trust network, discussed in Section \ref{realWorld}, are shared and discussed. For each real-world network, the results are compared to results from network models with similar average degree, clustering or modularity. We focus on the formation of echo chambers. We study balanced starting conditions and we introduce a fraction of completely stubborn actors. For both real-world networks, the REC and PR filtering algorithms are compared.

@@

The formation of echo chambers versus the fraction of stubborn actors, $f_s$, in the Last.fm on-line social network is studied. Results are compared to results from network models, all with a size of $N = 8003$ nodes: (i) WS model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM with 53 communities of size 151 nodes each, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) SBM-WS with 53 communities of size 151 nodes each, each community has WS parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. Results for the PR and REC filtering algorithms are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The SBM-WS has similar clustering and modularity as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering or modularity that drives the formation of echo chambers. Other topologies must play a role as well. A difference between the real-world Last.fm network and the SBM-WS lies in the degree distribution: the Last.fm network has a heterogeneous instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to network models with similar clustering and modularity, but with a homogeneous degree distribution.\\ % add parameters of SBM-WS power??
\newline
It is interesting to notice that the Last.fm on-line social network coincides with the SBM with similar modularity. This may suggest that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of a similar size as the hampering due to a lack of clustering.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The Last.fm on-line social network is compared to network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The Last.fm on-line social network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line ($y=1$) represents a constant amount of echo chambers over time. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\newpage
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease. The increase is the strongest for network models with high clustering (WS, SBM-WS). Even though the Last.fm network has similar clustering as these models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. Results for the Last.fm network coincide with those for the SBM. This seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering. \\
\newline
Another difference, beside the degree distribution, between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The SBM and SBM-WS have communities of a single, fixed size. This difference in community structure does not play a noticeable role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_REC.png}
    \label{pr_pow_single}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_PR.png}
    \label{rec_pow_single}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. The SBM-WS with heterogeneous community sizes is compared to the SBM-WS with a single community size. Results are compared to the real-world Last.fm network with similar clustering and modularity. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. The SBM-WS with heterogeneous community sizes (SBM-WS1) is compared to the SBM-WS with a single community size (SBM-WS2). Results are compared to the real-world Last.fm network with similar clustering and modularity. The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line ($y=1$) represents a constant amount of echo chambers over time. Each curve is the average of $50$ independent simulations. Error bars are omitted for improved visualization.}
\label{power_vs_single}
\end{figure}

@@

Results for the PGP trust network are compared to results for network models with similar average degree, clustering or modularity. These models are, all with a size of $N = 10680$ nodes: (i) WS model with $\left<k\right> = 4$, $\beta = 0.2$ and $\left<cc\right> = 0.266 \pm 0.003$; (ii) SBM with 60 communities of size 178 nodes each, $p_\text{cl} = 0.02$, $p_\text{add} = 0.00005$, $\left<k\right> = 4.06 \pm 0.02$ and $Q = 0.868 \pm 0.002$; (iii) SBM-WS with 60 communities of size 178 nodes each, each community has WS parameters: $\left<k\right> = 4$, $\beta = 0.12$; edges are added between communities with probability $p_\text{add} = 0.00007$, $\left<k\right> = 4.74 \pm 0.01$, $\left<cc\right> = 0.263 \pm 0.003$ and $Q = 0.842 \pm 0.002$.\\
\newline
In Fig. \ref{echo_vs_fracRes_real_PGP} the formation of echo chambers versus the fraction of stubborn actors, $f_s$, is studied for the real-world PGP trust network and for the network models. This is done for both the PR and REC filtering algorithms. \\
\newline
The PR filtering algorithm (Fig. \ref{pr_real_PGP}) shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors for both the PGP trust network and the network models. The real-world PGP network has a smaller formation of echo chambers compared to network models with similar clustering and modularity (SBM-WS). This is most likely due to the heterogeneous degree pattern of the PGP network.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_PGP_8x7.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_PGP_8x7.png}
    \label{rec_real_PGP}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The PGP trust network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The PGP trust network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line ($y=1$) represents a constant amount of echo chambers over time. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real_PGP}
\end{figure}
\newpage
\noindent
The REC filtering algorithm (Fig. \ref{rec_real_PGP}) shows an increase in the formation of echo chambers with an increasing fraction of stubborn actors, followed by a decrease if the fraction of stubborn actors is further increased. This is visible for both the PGP trust network and the network models. The PGP network has a smaller formation of echo chambers than the network model with similar clustering and modularity (SBM-WS), due to its heterogeneous degree distribution. The data points for the SBM and the PGP trust network coincide, which may suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering.

@@

We made use of simple toy models of social media platforms to generate results regarding the evolution of the prevalence of the opinions and formation of echo chambers. These toy models consist of two layers, a network layer and an opinion dynamics layer. Filtering algorithms were implemented to replicate and simulate the content curation used by social media companies. Furthermore, stubbornness was introduced to study the interplay between individual resistance to change, network structure and filtering algorithm. For simplicity we only allowed two competing opinions (A and B). We studied two starting conditions: balanced and unbalanced starting conditions.\\

@@

When introducing stubbornness, we found that the interplay between the PR filtering algorithm and individual resistance resulted in a decrease in the formation of echo chambers for both the high clustering (WS, SBM-WS) and high modularity networks (SBM). The REF and REC filtering algorithms showed an increase in the formation of echo chambers with increasing stubbornness, followed by a decrease if stubbornness was further increased. This is mainly seen in networks with high clustering, however a small bump can also be observed in networks with high modularity. This interplay between the PR filtering algorithm and individual resistance versus the interplay between the REF and REC filtering algorithms and stubbornness is seen for both balanced and unbalanced starting conditions. Furthermore, it was found that including stubborn actors hampers the increase in the majority opinion (and thus prevents the formation of a final consensus state) in case of unbalanced starting conditions combined with the PR filtering algorithm.\\

@@

These results give us insight in the interplay between network structure, filtering algorithms and individual resistance. These insights may be crucial to understand phenomena such as polarization in society and to determine the impact of social media on these phenomena. They learn us that even though semantic filtering (PR) enhances the formation of echo chambers in the absence of stubborn actors, this enhancement does not stand on its own. An interplay between networks with high clustering or community structure is needed to effectively see the increase in echo chambers induced by semantic filtering. Random networks, combined with semantic filtering, do not enhance the formation of echo chambers. If there are no stubborn actors, recent or random filtering (REC or REF) seem to be more protective against the formation of echo chambers than semantic filtering. If there is a small fraction of stubborn actors, this scenario changes: semantic filtering starts to hamper the formation of echo chambers, whereas recent or random filtering begins to enhance polarization. It is often assumed that algorithmic personalization, tailored to ones preferences (e.g. semantic filtering), is one of the reasons for the appearance of echo chambers on social media \cite{Ge2020}\cite{Mohseni2018}\cite{Stark2020}. Our results may, however, suggest, that if there is stubbornness, heterogeneous exposure (REF or REC) to different opinions may be more harmful than homogeneous exposure (PR). Bail et al. found experimental results that may agree with this finding \cite{Bail2018}. This is a result that should be studied more thoroughly.\\

@@

It was found that the initial distribution of the opinions over the communities in networks with community structure (SBM) makes a difference. We investigated a scenario in which a fraction of the communities was given a single opinion, while the remaining communities were given an initial $50/50$ opinion distribution. This scenario was then compared to the scenario in which the total opinion distribution was randomly distributed. This was done for both a high and low modularity configuration and for both the REC and PR filtering algorithms. \\

@@

The REC filtering algorithm did not show a difference in the evolution of the prevalence of the opinions between the community distributed and randomly distributed scenario. For the PR filtering algorithm, it was found that the high modularity configuration hampers the increase in the majority opinion in the community distributed scenario compared to the randomly distributed scenario. For the low modularity network, such a discrepancy was not found. Furthermore, it was found that, when using the REC filtering algorithm, the system evolves towards a configuration with more mixed opinion neighborhoods in the community distributed scenario. This reduces polarization and was found for both the high and low modularity configuration. The randomly distributed scenario does not evolve towards mixed opinion neighborhoods, instead a small increase in the formation of echo chambers around the minority opinion is observed. Hence, polarization increases in this scenario.\\

@@

For the PR filtering algorithm, the randomly distributed scenario resulted in an increase in the formation of echo chambers around the majority opinion (for both the low and high modularity configuration). The community distributed scenario for the low modularity configuration hampers this increase. It was found that this is more than a result of the increased number of initial opinion A echo chambers due to the single opinion communities. For the high modularity network, the community distributed scenario showed an even larger hampering in the formation of echo chambers around the majority opinion. This is probably the result of: (i) an even higher initial number of echo chambers around the majority opinion than for the low modularity network; (ii) the hampering in the increase in the majority opinion that occurs for the high modularity, community distributed scenario. \\

@@

Finally, results for real-world networks were compared to those from network models. It was found that the formation of echo chambers in real-world networks is smaller than those in network models with similar clustering. The reason for this hampered formation of echo chambers was attributed to the heterogeneous degree pattern of the real-world networks. This hampering of echo chambers due to a heterogeneous degree distribution was also found by Perra and Rocha \cite{Perra2019} when they compared network models with heterogeneous degree distributions to models with homogeneous contact patterns.\\

@@

The hampering in the increase of the majority opinion in case of unbalanced starting conditions combined with the introduction of stubborn actors for the PR filtering algorithm agrees with related research on continuous opinion dynamics models combined with stubborn actors \cite{Como2013}. In this research, Acemo\u{g}lu et al. showed that introducing stubborn actors (with different opinions) in an inhomogeneous stochastic gossip process of continuous opinion dynamics leads to a situation in which consensus can no longer be reached \cite{Como2013}. This result is also found in the majority opinion dynamics model and in the voter model combined with stubborn actors (equally distributed over the two opinions) \cite{Galam2007}\cite{Mobilia2007}\cite{Yildiz2013}. In our case, the possible consensus around the majority opinion, that could have been reached when combining the PR filtering algorithm with unbalanced starting conditions, can no longer be obtained when introducing stubborn actors.\\

@@

It must be emphasized that there are limitations to our model and methods. Real-life social behavior and opinion dynamics on social media may be far from simple. As a consequence, there may be other dynamics of real-life opinion formation that are not well-captured by our model. More empirical data is therefore needed to compare the results, obtained from our model, against \cite{Castellano2009}\cite{Pawel}. Other limitations may be the small system sizes that were used in our simulations ($N = 10^3$). It is unclear how system size affects the results and it is recommended that the model is tested on both larger network models and real-world networks. Limitations due to the activation mechanism, may be that we used a constant activation probability. Analyses of real social networks show that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. It might be an interesting extension to switch to the heterogeneous activation probabilities and investigate possible impacts on the dynamics. Another possible shortcoming is the binary nature of the model. There are real-life cases in which people have to choose between two competing opinions. These situations, however, do not cover the complete range of opinion dynamics situations. It would be interesting to run the model on configurations where more than two opinions are allowed or to switch to continuous opinion dynamics. We also did not differentiate between the actors in our model, except for non-stubborns versus stubborns. This combined with the homogeneous degree distribution of our network models may be another limitation. Real-life social media platforms display a heterogeneous contact pattern, where the `hubs' are often more influential than other users, not only by their larger reach, but also by their status. This may impact the dynamics of opinions \cite{Fan2020}\cite{Wang2019}. We could adapt our model to mimic this social media behavior. Therefore, a heterogeneous network model should be introduced, combined with an influence parameter. This parameter is used to determine the influence of an actor on others: the opinion of a more influential actor is easier adopted by the others. In the model, the high degree nodes should then be made more influential. The fixed stubbornness of the actors might be another shortcoming. Even though the stubbornness parameter $r$ could be individualized to each actor, we did not include the possibility to be more stubborn towards opinions coming from one friend than from another. This idea may be linked to the influence of certain friends: somebody may adopt another viewpoint easier from certain friends than from others.\\
